ProbabilityBoundsAnalysis.jl:\\ an arithmetic of probabilities and sets of probabilities

### 2021-09-13 paper/paper.tex
adds fig2 and continues section 2

Probability Bounds Analysis combines interval arithmetic with probability theory, and provides a representation of sets of distributions in structures called probability boxes (p-boxes), which generalise both distribution functions and intervals. P-boxes generally return interval bounds on all probabilistic quantities, for example samples, cdfs, and probability measures are all intervals. This framework also allows for the comprehensive propagation of probabilities through calculations in a rigorous way, in a similar fashion that interval arithmetic does for sets. As such, ProbabilityBoundsAnalysis.jl gives a rigorous arithmetic of random variables, where both marginal (univariate) and dependency information can be known, partially known or missing completely.

@@

\noindent Indeed it was Kolmogorov who originally asked what the result of a sum of two distributions without knowing their joint distribution. This was answered for the sum by Marakov \cite{makarov1982estimates}, who showed the result was a set of distributions and was able to provide bounds on this function. Sklar, Schweizer and Frank generalised this result to other (positive) binary operations \cite{frank1987best,schweizer2011probabilistic}. In this pursuit they created copulas, a general way to encode probabilistic dependence independently from marginals, and a now essential object used in probabilistic modelling. In his seminal dissertation \cite{williamson1989probabilistic}, Williamson described an algorithm for efficiently performing these arithmetic operations, which can give guaranteed bounds on probability distributions in terms of an upper and lower cdf. He called his method \textit{Probabilistic Arithmetic} and his sets of distributions \textit{Dependency Bounds}. Since then the method has been generalised \cite{ferson2015constructing,ferson1996whereof,ferson2004arithmetic} to most of the base binary and unary operations that would be present in a programming language. Probability Boxes (p-boxes) are the name now given to these structures, and Probability Bounds Analysis (PBA) the name of the method.

@@

where the max operator is required when $\underline{F}(b) < \overline{F}(a)$. Note the same can be achieved by using the standard formula for finding the measure from the cdf, $F(b)- F(a)$, and using interval arithmetic.

@@

  \caption{A precise distribution, an interval and a probability box}
  \label{fig:figure3}

@@

\noindent That is, if the set $U$ does not intersect the interval $X$, $\mathbb{P}_{X}=0$, i.e. $U$ certainly does not contain any of the random variables. If the X is fully contained in $U$, $\mathbb{P}_{X}=1$, i.e. $U$ certainly contains all of the random variables. And finally if $U$ intersects (but does not contain) $X$, the probability measure is the vacuous probability interval $\mathbb{P}_{X} = [0, 1]$, i.e. we are completely uncertain about the containment. Note that the $P_{X} = \{0, 1, [0,1]\}$ is due to the interval bounds being step function. Generally p-boxes may yield any probability interval.

@@

>>>>

Probability Bounds Analysis combines interval arithmetic with probability theory, and provides a representation of sets of distributions in structures called probability boxes (p-boxes), which generalise both distribution functions and intervals. P-boxes generally return interval bounds on all probabilistic quantities, for example samples, cdfs, and probability measures are all intervals. This framework also allows for the comprehensive propagation of probabilities through calculations in a rigorous way, in a similar fashion that interval arithmetic does for sets. As such, ProbabilityBoundsAnalysis.jl gives a rigorous arithmetic of random variables, where both marginal (univariate) and dependency information can be known, partially known or missing completely. We describe the main theoretical elements of probability bounds analysis, and describe a minimal implementation of the method and it's main algorithms in Julia.

@@

\noindent Indeed it was Kolmogorov who originally asked about the result of a sum of two distributions without knowing their joint distribution. This was answered for the sum by Marakov \cite{makarov1982estimates}, who showed the result was a set of distributions and was able to provide bounds on this function. Sklar, Schweizer and Frank generalised this result to other (positive) binary operations \cite{frank1987best,schweizer2011probabilistic}. In this pursuit they created copulas, a general way to encode probabilistic dependence independently from marginals, and a now essential object used in probabilistic modelling. In his seminal dissertation \cite{williamson1989probabilistic}, Williamson described an algorithm for efficiently performing these arithmetic operations, which can give guaranteed bounds on probability distributions in terms of an upper and lower cdf. He called his method \textit{Probabilistic Arithmetic} and his sets of distributions \textit{Dependency Bounds}. Since then the method has been generalised \cite{ferson2015constructing,ferson1996whereof,ferson2004arithmetic} to most of the base binary and unary operations that would be present in a programming language. Probability Boxes (p-boxes) are the name now given to these structures, and Probability Bounds Analysis (PBA) the name of the method.

@@

where the max operator is required when $\underline{F}(b) < \overline{F}(a)$. Note the same can be achieved by using the standard formula for finding the measure from the cdf, $\mathbb{P}(U) = F(b)- F(a)$, and using interval arithmetic.

@@

  \caption{A precise distribution, an interval and a probability box}
  \label{fig:figure1}

@@

\noindent That is, if the set $U$ does not intersect the interval $X$, $\mathbb{P}_{X}=0$, i.e. $U$ certainly does not contain any of the random variables. If $X$ is fully contained in $U$, $\mathbb{P}_{X}=1$, i.e. $U$ certainly contains all of the random variables. And finally if $U$ intersects (but does not contain) $X$, the probability measure is the vacuous probability interval $\mathbb{P}_{X} = [0, 1]$, i.e. we are completely uncertain about the containment. Note that the $P_{X} = \{0, 1, [0,1]\}$ is due to the interval bounds being step functions. Generally p-boxes may yield any probability interval.

@@

We follow the representation first introduced by Williamson and Downs \cite{williamson1990probabilistic}, where a discrete upper and lower approximations of distribution functions are constructed using the inverse cdfs. A minimal p-box structure is as follows:

\begin{lstlisting}[language = Julia]
  using IntervalArithmetic

  struct pbox{T<: Real}
    u :: Vector{T}    # right descrete inverse
    d :: Vector{T}    # left descrete inverse
    m :: Interval{T}  # interval mean
    v :: Interval{T}  # interval variance
    shape :: String   # shape
  end
\end{lstlisting}

\noindent where the vectors $u$ and $d$ are the "up" and "down", or the "left" and "right", edges of the p-box. The intervals $m$ and $v$ are the interval moments, and the $shape$ is a string containing any distribution family information.

  \centering
  \includegraphics[width=.3\textwidth]{../examples/JuliaCon/fig2/fig2_pbox1.pdf}\hfill
  \includegraphics[width=.3\textwidth]{../examples/JuliaCon/fig2/fig2_pbox2.pdf}\hfill
  \includegraphics[width=.3\textwidth]{../examples/JuliaCon/fig2/fig2_pbox3.pdf}

  \caption{A precise distribution, an interval and a probability box}
  \label{fig:figure2}
